LabMate: supporting types for MATLAB

### 2024-03-28 writing/IMEKOWC2024/LabMate/paper.tex
[ writing ] fix minor typos

To rectify this situation, we have developed LabMate, which is a tool
to reify current virtuous engineering practices as a formal language
of MATLAB comments. MATLAB remains in the dark, but LabMate reads,
assesses, and transforms MATLAB programs in accordance with these
comments.
%
Behind the scenes, LabMate is retrofitting an expressive type system
to MATLAB, with the more of the meaning of programs recorded in their
types.
%
This gives a lightweight and low-cost way for MATLAB programmers to
express their intent, and in a language they are already working in,
rather than starting over from scratch.
%
While type systems and their theory is a well established discipline
of computer science, the development of LabMate has a several novel
advances novel focus on the algebraic structure required to classify
matrices and the meanings of the quantities therein, e.g., their
physical dimensions. LabMate brings advanced type systems to
effectiveness within, rather than instead of, existing scientific and
engineering toolchains and practices.

@@

Under the hood, LabMate translates MATLAB expressions to typed
expressions in its own core type theory.
%
Since matrices feature heavily in MATLAB code, the type of matrices
play a central role in this type theory.
%
We make this type rather precise, by allowing the type of each entry
in the matrix to vary according to the row and column of the entry:
%
we parameterise the type of matrices by a type $R$ associated with
each row, a type $C$ associated with each column, and a type $E(x,y)$
which might vary depending on the values of parameters $x : R$ and
$y : C$, as well as two lists $rs : \List{R}$ and $\List{C}$ --- the
idea being that the $i$th entry in the list $rs$ and the $j$th entry
in the list $cs$ determines the type of the $(i, j)$ coordinate in the
matrix.
%
Summarised as an inference rule, matrix types are thus formed as follows:
%
\[
  \infer{\istype{\Matrix{R}{C}{E}{rs}{cs}}}
  {
    \istype{R}
    &\quad
    \istype{C}
    &\quad
    x : R, y : C \vdash \istype{E(x,y)}
    &\quad
    rs : \List{R}
    &\quad
    cs : \List{C}
  }
\]

>>>>

To rectify this situation, we have developed LabMate, which is a tool
to reify current virtuous engineering practices as a formal language
of MATLAB comments. MATLAB remains in the dark, but LabMate reads,
assesses, and transforms MATLAB programs in accordance with these
comments.
%
Behind the scenes, LabMate is retrofitting an expressive type system
to MATLAB, with the more of the meaning of programs recorded in their
types.
%
This gives a lightweight and low-cost way for MATLAB programmers to
express their intent, and in a language they are already working in,
rather than starting over from scratch.
%
While type systems and their theory is a well established discipline
of computer science, the development of LabMate has a several novel
advances on the algebraic structure required to classify
matrices and the meanings of the quantities therein, e.g., their
physical dimensions. LabMate brings advanced type systems to
effectiveness within, rather than instead of, existing scientific and
engineering toolchains and practices.

@@

Under the hood, LabMate translates MATLAB expressions to typed
expressions in its own core type theory.
%
Since matrices feature heavily in MATLAB code, the type of matrices
plays a central role in this type theory.
%
We make this type rather precise, by allowing the type of each entry
in the matrix to vary according to the row and column of the entry:
%
we parameterise the type of matrices by a type $R$ associated with
each row, a type $C$ associated with each column, and a type $E(x,y)$
which might vary depending on the values of parameters $x : R$ and
$y : C$, as well as two lists $rs : \List{R}$ and $\List{C}$ --- the
idea being that the $i$th entry in the list $rs$ and the $j$th entry
in the list $cs$ determines the type of the $(i, j)$ coordinate in the
matrix.
%
Summarised as an inference rule, matrix types are thus formed as follows:
%
\[
  \infer{\istype{\Matrix{R}{C}{E}{rs}{cs}}}
  {
    \istype{R}
    &\quad
    \istype{C}
    &\quad
    x : R, y : C \vdash \istype{E(x,y)}
    &\quad
    rs : \List{R}
    &\quad
    cs : \List{C}
  }
\]
